---
title: "Apoptosis_ROS_Article"
author: "Ricardo Oliveira"
date: "2024-11-13"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Script for Somatic Mutation Analysis of Apoptosis and Reactive Oxygen Species Defense Genes


#### This script presents the methodology and results of my analysis of somatic mutations in genes involved in apoptosis and defense against reactive oxygen species (ROS). The following sections contain detailed steps and code used to identify, filter, and analyze mutations in these key gene sets, providing insights into their potential roles and associations with cancer progression and therapeutic responses.

1. **Filtering of Mutations by Cancer Type**  
   We begin by filtering somatic mutations across four types of cancer: breast, prostate, esophagogastric, and colorectal. The datasets for these cancers were sourced from the CBioPortal database. Notably, these databases were selected randomly and represent collections that have not been extensively detailed in the literature.

2. **Initial Data Handling in Microsoft Excel**  
   After downloading the datasets, preliminary filtering was done in Microsoft Excel:
   - Removed duplicate mutations identified within samples from the same individual.
   - Ensured each sample reflected unique mutation data for accurate downstream analyses.

3. **Data Import into RStudio Environment**  
   Following Excel processing, the filtered data were imported into the RStudio environment to prepare for further analysis and visualization.

4. **Common Gene Analysis in R**  
   In R, the analysis focuses on identifying common target genes with somatic mutations across the cancer types. This step checks for overlapping genes among cancers rather than investigating identical mutations. The objective is to determine which mutated genes are consistently present in multiple cancer types, highlighting potential cross-cancer gene targets for further study. 

--- 

This structure sets up a clear pathway from data acquisition to the analysis focus in R.

### Necessary Packages
```{r}
library(tidyverse)
library(ggplot2)
library(GOplot)
library(RColorBrewer)
library(dplyr)
library(UpSetR)
library(rJava)
library(UpSetR)
library(grid)
library(ggforce)
library(clusterProfiler)
library(org.Hs.eg.db)

```

### Here, an Upset plot visualization is being created to examine the common genes shared across different cancer types. Following this, an Over-Representation Analysis (ORA) will be conducted to assess the functional enrichment of these genes. This enrichment analysis aims to uncover the biological functions and pathways in which these shared genes are involved, providing insights into their roles in cancer across multiple tissue types.

#### It is important to note that before this, the file with the mutations in Apoptosis and ROS defense genes was imported from the code:

#### ros_apoptose_tudo <- read.delim(“~/tabela completa (ROS e APOP) - TABELA COMPLETA.tsv”, header=TRUE)

```{r}
# Inserting the general table
ros_apoptose_tudom <- read.delim("C:/Users/ricar/Downloads/tabela completa (ROS e APOP) - TABELA COMPLETA.tsv", header=FALSE)

# Taking the gene symbols and cancer type columns into a new data.frame
ros_apoptose_tudo <- NULL

ros_apoptose_tudo$Hugo_Symbol <- ros_apoptose_tudom$V1

ros_apoptose_tudo$Tipo <- ros_apoptose_tudom$V15

ros_apoptose_tudo$Grupo <- ros_apoptose_tudom$V16

ros_apoptose_tudo <- as.data.frame(ros_apoptose_tudo)


# Creating gene lists based on cancer type from the column "Tipo" in 'ros_apoptose_tudo'
# "Hugo_Symbol" contains gene names; "Tipo" specifies cancer type
mama_mateus_ros_apo <- ros_apoptose_tudo$Hugo_Symbol[ros_apoptose_tudo$Tipo == "Mama"]
prostata_mateus_ros_apo <- ros_apoptose_tudo$Hugo_Symbol[ros_apoptose_tudo$Tipo == "Prostata"]
esofagog_mateus_ros_apo <- ros_apoptose_tudo$Hugo_Symbol[ros_apoptose_tudo$Tipo == "Esofagogastrico"]
colorretal_mateus_ros_apo <- ros_apoptose_tudo$Hugo_Symbol[ros_apoptose_tudo$Tipo == "Colorretal"]


# Creating a combined list with gene names for each cancer type
lista_apop_ros <- list("Breast" = mama_mateus_ros_apo, 
                       "Prostate" = prostata_mateus_ros_apo,
                       "Esophagogastric" = esofagog_mateus_ros_apo,
                       "Colorectal" = colorretal_mateus_ros_apo)

# Generating an Upset plot to visualize gene overlaps across cancer types
upset(fromList(lista_apop_ros), order.by = "freq")

# Creating gene lists to find unique or common genes across different cancer types
list_pivic_ros_apo_mama <- setdiff(mama_mateus_ros_apo, c(prostata_mateus_ros_apo, esofagog_mateus_ros_apo, colorretal_mateus_ros_apo))
list_pivic_ros_apo_prostata <- setdiff(prostata_mateus_ros_apo, c(mama_mateus_ros_apo, esofagog_mateus_ros_apo, colorretal_mateus_ros_apo))
list_pivic_ros_apo_prostata_mama <- Reduce(intersect, list(prostata_mateus_ros_apo, mama_mateus_ros_apo))

# Finding genes common to all four cancer types
list_genes_part_final <- Reduce(intersect, list(prostata_mateus_ros_apo, mama_mateus_ros_apo, colorretal_mateus_ros_apo, esofagog_mateus_ros_apo))
list_genes_CM_CP_2_apop_ros <- setdiff(mama_mateus_ros_apo, prostata_mateus_ros_apo)

# Performing enrichment analysis (ORA) on common genes
enriqueciment_genes_pivic_tudo <- enrichGO(
  gene = list_genes_part_final,
  OrgDb = "org.Hs.eg.db",
  ont = "BP",
  pAdjustMethod = "BH", 
  keyType = 'SYMBOL')

# Plotting enrichment results and saving as PNG
barplot(enriqueciment_genes_pivic_tudo)
ggsave("my_plot.png", width = 15, height = 10, dpi = 600)

# Visualizing top enriched terms in a custom barplot
enriqueciment_genes_pivic_tudo <- enriqueciment_genes_pivic_tudo[order(enriqueciment_genes_pivic_tudo$p.adjust), ]
list_genes_todos <- head(enriqueciment_genes_pivic_tudo, 8)

list_genes_todos

p1 <- ggplot(list_genes_todos, aes(x = reorder(Description, p.adjust), y = -log10(p.adjust))) +
  geom_bar(stat = "identity", aes(fill = -log10(p.adjust)), color = "black") +
  scale_fill_gradient(low = "lightpink", high = "darkblue") +
  coord_flip() +
  labs(x = "", y = "-log10(p.adjust)",
       title = "Enrichment of Biological Processes from common genes found in the four types of cancer",
       caption = "Fonte: Santana et al., 2023") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 12, colour = "black"),
        plot.title = element_text(size = 12, hjust = 2, colour = "black"),
        plot.caption = element_text(size = 10, hjust = 0),
        axis.text.x = element_text(angle = 45, hjust = 0, colour = "black"))

# Display plot
print(p1)

# Enrichment analysis for genes common in Breast and Prostate cancers
list_genes_CM_CP_ros_apop <- Reduce(intersect, list(mama_mateus_ros_apo, prostata_mateus_ros_apo))

list_genes_CM_CP_ros_apop

enriqueciment_genes_CM_CP <- enrichGO(
  gene = list_genes_CM_CP_ros_apop,
  OrgDb = "org.Hs.eg.db",
  ont = "BP",
  pAdjustMethod = "BH",
  keyType = 'SYMBOL')

# Display enrichment barplot
barplot(enriqueciment_genes_CM_CP)

# Order results by adjusted p-value and select the top 8 enriched terms
enriqueciment_genes_CM_CP <- enriqueciment_genes_CM_CP[order(enriqueciment_genes_CM_CP$p.adjust), ]
list_genes_CM_CP <- head(enriqueciment_genes_CM_CP, 8)

# Check top terms
print(list_genes_CM_CP)

# Plot top enriched terms in a custom bar plot
p2 <- ggplot(list_genes_CM_CP, aes(x = reorder(Description, p.adjust), y = -log10(p.adjust))) +
  geom_bar(stat = "identity", aes(fill = -log10(p.adjust)), color = "black") +
  scale_fill_gradient(low = "lightpink", high = "darkblue") +
  coord_flip() +
  labs(
    x = "", y = "-log10(p.adjust)",
    title = "Enrichment of Biological Processes from common genes in Breast and Prostate cancers",
    caption = "Fonte: Santana et al., 2023"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 12, colour = "black"),
    plot.title = element_text(size = 12, hjust = 0.5, colour = "black"),
    plot.caption = element_text(size = 10, hjust = 0),
    axis.text.x = element_text(angle = 45, hjust = 1, colour = "black")
  )

# Display plot
print(p2)

```







### In this section, a filtering process was applied to isolate specific genetic variants for submission to EnsemblVEP. The purpose of this submission is to obtain pathogenicity predictions for these variants, as well as to identify any prior literature documentation on these mutations. This approach helps determine the potential impact of the variants in terms of disease association and clinical relevance by cross-referencing them with known databases and research publications


```{r}
### Loading Data Frames ###

# Breast cancer datasets
brca_1 <- read.delim("C:/Users/ricar/Downloads/teste_teste/breast_ink4_msk_2021/brca_mbcproject_2022/data_mutations.txt")
brca_2 <- read.delim("C:/Users/ricar/Downloads/teste_teste/breast_ink4_msk_2021/breast_ink4_msk_2021/data_mutations.txt", comment.char="#")
brca_3 <- read.delim("C:/Users/ricar/Downloads/teste_teste/breast_ink4_msk_2021/brca_igr_2015/data_mutations.txt")

# Colorectal cancer datasets
crc_1 <- read.delim("C:/Users/ricar/Downloads/crc_msk_2017/crc_msk_2017/data_mutations.txt")
crc_2 <- read.delim("C:/Users/ricar/Downloads/teste_teste/breast_ink4_msk_2021/crc_nigerian_2020/data_mutations.txt", comment.char="#")

# Prostate cancer datasets
prostate_1 <- read.delim("C:/Users/ricar/Downloads/teste_teste/breast_ink4_msk_2021/prad_su2c_2019/data_mutations.txt")
prostate_2 <- read.delim("C:/Users/ricar/Downloads/teste_teste/breast_ink4_msk_2021/prad_mcspc_mskcc_2020/data_mutations.txt", comment.char="#")
prostate_3 <- read.delim("C:/Users/ricar/Downloads/teste_teste/breast_ink4_msk_2021/mpcproject_broad_2021/data_mutations.txt")
prosstate_4 <- read.delim("C:/Users/ricar/Downloads/teste_teste/breast_ink4_msk_2021/prad_mich/data_mutations.txt", comment.char="#")

# Esophagogastric cancer dataset
esg_1 <- read.delim("C:/Users/ricar/Downloads/teste_teste/breast_ink4_msk_2021/egc_msk_2017/data_mutations.txt")

#### Filtering Data ####

# Load a dataset of gene functions for filtering
human_gene_function <- read.delim2("~/Mestrado_Rstudio_org/human_gene_function.txt")

# Filter rows with "Apoptosis" and "ROS Defense" processes
genes_pivic <- subset(human_gene_function, mito_process %in% c("Apoptosis", "ROS Defense"))

# Create a vector with gene symbols
genes_pivic <- genes_pivic$HGNC_symbol

# Filter each cancer dataset to include only the specified genes
brca_1 <- subset(brca_1, Hugo_Symbol %in% genes_pivic)
brca_2 <- subset(brca_2, Hugo_Symbol %in% genes_pivic)
brca_3 <- subset(brca_3, Hugo_Symbol %in% genes_pivic)
crc_1 <- subset(crc_1, Hugo_Symbol %in% genes_pivic)
crc_2 <- subset(crc_2, Hugo_Symbol %in% genes_pivic)
prostate_1 <- subset(prostate_1, Hugo_Symbol %in% genes_pivic)
prostate_2 <- subset(prostate_2, Hugo_Symbol %in% genes_pivic)
prostate_3 <- subset(prostate_3, Hugo_Symbol %in% genes_pivic)
prosstate_4 <- subset(prosstate_4, Hugo_Symbol %in% genes_pivic)
esg_1 <- subset(esg_1, Hugo_Symbol %in% genes_pivic)

# Combine the tables for analysis
library(tidyverse)

# Selecting the HGVSc column for variant information
brca_1_a <- brca_1$HGVSc
brca_2_a <- brca_2$HGVSc
brca_3_a <- brca_3$HGVSc
crc_1_a <- crc_1$HGVSc
crc_2_a <- crc_2$HGVSc
prostate_1_a <- prostate_1$HGVSc
prostate_2_a <- prostate_2$HGVSc
prostate_3_a <- prostate_3$HGVSc
prosstate_4_a <- prosstate_4$HGVSc
esg_1_a <- esg_1$HGVSc

# Combine all variant information into a single data frame
tabela_variant <- c(brca_1_a, brca_2_a, brca_3_a, crc_1_a, crc_2_a, prostate_1_a, prostate_2_a, prostate_3_a, prosstate_4_a, esg_1_a)
jss <- data.frame(Valores = tabela_variant)

# Export the table of variants to a text file
write.table(jss, "variants_pivic.txt", quote = F, row.names = F)

# Find unique values of variants
unique(jss$Valores)

# Selecting the Transcript_ID column for further analysis
brca_1_a <- brca_1$Transcript_ID
brca_2_a <- brca_2$Transcript_ID
brca_3_a <- brca_3$Transcript_ID
crc_1_a <- crc_1$Transcript_ID
crc_2_a <- crc_2$Transcript_ID
prostate_1_a <- prostate_1$Transcript_ID
prostate_2_a <- prostate_2$Transcript_ID
prostate_3_a <- prostate_3$Transcript_ID
prosstate_4_a <- prosstate_4$Transcript_ID
esg_1_a <- esg_1$Transcript_ID

# Combine Transcript_IDs for further filtering
tabela_variant <- c(brca_1_a, brca_2_a, brca_3_a, crc_1_a, crc_2_a, prostate_1_a, prostate_2_a, prostate_3_a, prosstate_4_a, esg_1_a)

# Filter data using a variable in the environment (not yet defined in script)
variantes <- subset(LAJxPp8nhMiBN6bV, Feature %in% tabela_variant)

# Clean up the 'Feature' column by removing any text after a period
LAJxPp8nhMiBN6bV$Feature <- gsub("\\..*$", "", LAJxPp8nhMiBN6bV$Feature)

# Selecting specific columns to create a new data frame
novo_dados <- variantes[, c("X.Uploaded_variation", "SYMBOL", "PUBMED")]

# Display the new data frame
print(novo_dados)
```